sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-04 doc/paper-jss/paper-jss.Rmd
Edit

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station'
    address: |
      | Pacific Biological Station, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center \AND'
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    address: ''
    affiliation: 'Pacific Biological Station'
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center'
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are pervasive and available in increasingly large datasets; however, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or Bayesian inference via \pkg{tmbstan} and \pkg{rstan}). We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations). Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation. Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be 4–16 times faster than INLA. We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
keywords:
  # at least one keyword must be supplied
  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{lscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
  \usepackage{pifont}
  \usepackage{newunicodechar}
  \newunicodechar{✓}{\ding{51}}
  \newunicodechar{✗}{\ding{55}}
output: rticles::jss_article
---

>>>>

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station'
    address: |
      | Pacific Biological Station, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center \AND'
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    address: ''
    affiliation: 'Pacific Biological Station'
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center'
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are pervasive and available in increasingly large datasets; however, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}. We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations). Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation. Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be 4–16 times faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
keywords:
  # at least one keyword must be supplied
  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects model (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects model (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{lscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
  \usepackage{pifont}
  \usepackage{newunicodechar}
  \newunicodechar{✓}{\ding{51}}
  \newunicodechar{✗}{\ding{55}}
output: rticles::jss_article
---
